\ifdefined\XeTeXrevision\else\pdfoutput=1\fi
\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\graphicspath{{./}{figures/}{../figures/}}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage[inline]{enumitem}
\usepackage{xspace}

\newtheorem{definition}{Definition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{proposition}{Proposition}
\newcommand{\Dcont}{\Delta_{\mathrm{iso}}^{\mathrm{IRT}}\xspace}

\title{Isomorph-Eval: Structurally Equivalent Benchmarks Reveal\\
That Annotation-Chain Replay Produces False Contamination Signals}

\author{Jung Min Kang\thanks{Corresponding author. Email: gangjeongmin23@gmail.com}\\
Independent Researcher\\
Seoul, South Korea}

\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
We present \textbf{Isomorph-Eval}, a framework for generating
structurally equivalent benchmark variants via \emph{$\tau$-isomorphism},
a graph-theoretic definition that preserves operation types, dependency
structure, and complexity weights while ensuring complete surface
novelty. The framework's four-stage pipeline (Parser $\to$ Mutator
$\to$ Generator $\to$ Verifier) produces unlimited verified variants
of existing benchmark items with automatically computed answers.

Applying Isomorph-Eval to GSM8K, we discover that
\emph{annotation-chain replay}---the standard approach to computing
variant answers by substituting values into the original solution
chain---produces systematically wrong answers in 41\% of items. We
identify eight bug classes (value conflation, phantom leaf values,
positional replacement collisions, magnitude inflation, answer
invariance, partial name replacement, node reuse, and inconsistent
entity replacement) that create
false contamination signals: all five evaluated models appeared
uniformly contaminated ($\Delta \approx +0.42$) on the unverified
data, including a comparison model whose training data composition
has not been fully disclosed.

After independent verification and pipeline repair, the signal
collapses. On 63 entity-clean verified items, four of five models show equivalent
performance on originals and variants
($\Delta_{\mathrm{iso}} \in [+0.036, +0.047]$, all $p > 0.05$),
while Llama~3.1~8B shows a mild gap
($\Delta_{\mathrm{iso}} = +0.114$) consistent with known numeric
perturbation sensitivity. We argue that independent answer and entity
verification are necessary for any benchmark variant generation
framework, and release the complete pipeline as open-source software.
\end{abstract}

\input{paper_outline_and_intro}
\input{section4_methodology}
\input{section5_6_empirical}

\section{Conclusion}
\label{sec:conclusion}

We have presented Isomorph-Eval, a framework for generating verified
structurally equivalent benchmark variants grounded in a formal
definition of $\tau$-isomorphism. The framework makes four
contributions: (1)~$\tau$-isomorphism formalizes structural equivalence
via graph-theoretic invariants; (2)~the Isomorphic Engine generates
unlimited verified variants with automatically computed answers;
(3)~the identification of eight bug classes in annotation-chain replay
that produce false contamination signals; and (4)~empirical evidence
that on verified data, current LLMs show robust reasoning transfer to
novel numeric contexts.

The central lesson of this work is methodological: annotation-chain
replay---substituting values into the original solution chain to compute
variant answers---produced systematically wrong answers in 41\% of
items, creating an apparent contamination signal of
$\Delta \approx +0.42$ that was uniform across all models including a
comparison model. After independent verification, the signal collapsed
to $\Delta \approx +0.04$--$0.11$, with one model showing a mild gap
consistent with known perturbation sensitivity. This finding
has implications for any benchmark perturbation methodology:
\emph{answer verification must be independent of the generation
pipeline}.

We release the complete pipeline as open-source software at
\href{https://github.com/testofschool/isomorph-eval}{the project repository}.

\paragraph{Future work.}
Three directions are priorities: (1)~extending the variant generation
pipeline to handle the 31~items with irreducible structural issues
(e.g., multi-step phantom intermediates), further reducing entity-level
quarantine, and expanding to the full
GSM8K test set; (2)~extending the framework beyond mathematical
reasoning to logic, code generation, and scientific reasoning;
(3)~fitting full 2PL IRT models with $\geq 30$ models to enable
proper item parameter estimation and DIF testing.

\bibliographystyle{plainnat}
\bibliography{references}

\appendix
\section{Unverified Results}
\label{app:unverified}

For transparency, we report the results on all 100 items before
independent answer verification. These results should \textbf{not} be
used for contamination claims due to the data quality issues documented
in Section~\ref{sec:setup}.

\begin{table}[h]
\centering
\small
\caption{Results on all 100 items (before verification). These results are inflated by broken variant answers.}
\begin{tabular}{lcccc}
\toprule
Model & $N$ & Acc$_{\mathrm{orig}}$ & Acc$_{\mathrm{iso}}$ & $\Delta_{\mathrm{raw}}$ \\
\midrule
Llama 3.1 8B & 100 & 92.0\% & 50.3\% & $+0.417$ \\
Llama 4 Scout 17B & 100 & 94.0\% & 53.8\% & $+0.402$ \\
Qwen3 32B & 70 & 100.0\% & 53.5\% & $+0.465$ \\
GPT-OSS 120B & 20 & 100.0\% & 57.0\% & $+0.430$ \\
Llama 3.3 70B & 14 & 100.0\% & 55.7\% & $+0.443$ \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
