% ==========================================================================
% Sections 5-6: Experimental Setup and Results (Empirical) — v3
% ==========================================================================

\section{Experimental Setup}
\label{sec:setup}

\subsection{Evaluation Dataset}
We apply the Isomorphic Engine (Section~\ref{sec:engine}) to the
GSM8K test set~\cite{cobbe2021gsm8k}, parsing all 1{,}319 problems
via the rule-based annotation parser. After applying the strict
validation pipeline---filtering problems with fractional leaf values,
value-conflation between inputs and implicit constants, non-integer
variant answers, negative intermediates, excessively large answers
($>10^8$), and text-replacement mismatches---the final dataset contains
\textbf{667 original items with 3{,}199 verified $\tau$-isomorphic
variants} (3{,}866 evaluation entries total, $\sim$4.8 variants per
item). We evaluate a \textbf{stratified subset of 100 items} with
480 variants (580 evaluation entries per model), selected by uniform
sampling across answer-magnitude quantiles as a proxy for difficulty.
Each item is limited to 5~variants. All variant answers are
computed by replaying the original GSM8K solution annotations with
substituted input values, distinguishing question-derived inputs from
computation constants.

\paragraph{Dataset validation.}
Deterministic verification confirms 100\% positive-integer answers,
100\% structural fingerprint preservation ($\tau$-signature match),
and 100\% reasonable answer range. The strict validation pipeline
rejected 19{,}826 candidate variants (63.5\% non-integer answers,
12.1\% text-replacement mismatches, 9.6\% negative values, 1.2\%
negative intermediates, 0.7\% overflow), ensuring that surviving
variants produce coherent natural-language problems with verified
answers. A spot-check of 10 random items confirms high word overlap
(80--96\%) between original and variant questions, with differences
limited to numeric values and entity names.

\subsection{Models}
We evaluate three models spanning two model families and a range
of parameter counts:

\begin{itemize}[nosep, leftmargin=*]
\item \textbf{Llama~3.1 8B} (Meta, dense, 8B parameters)
\item \textbf{Llama~4 Scout 17B} (Meta, MoE, 17B active parameters)
\item \textbf{Qwen3 32B} (Alibaba, dense, 32B parameters)
\end{itemize}

\noindent All models are accessed via the Groq inference API at
temperature $T = 0.0$ (greedy decoding) with chain-of-thought prompting
and \verb|\boxed{}| answer extraction, with fallback to
\verb|####|-delimited and last-number patterns. Extraction rates
exceed 95\% for both Llama-family models (100\%/99\% for 8B,
96\%/95\% for Scout) and 100\%/89\% for Qwen3 (whose thinking-token
format occasionally produces malformed output). The extraction-rate
gap between original and variant items is $\leq$1.4\% for all models,
confirming that the contamination delta is not an artifact of
differential extraction. Qwen3 was evaluated on 42 of 100~items due
to API daily token limits; both Llama models were evaluated on all
100~items.

\subsection{Protocol}
Each model receives every original item and every variant item exactly
once ($T = 1$ trial per item, greedy decoding). We compute the
item-level contamination delta $\delta_i$ as the difference between
the model's accuracy on the original item and its mean accuracy across
the item's $\tau$-isomorphic variants. The model-level metrics
$\Delta_{\mathrm{raw}}$ and $\Dcont$ are computed as described in
Section~\ref{sec:delta}, with the IRT-weighted variant using
variance-based discrimination weights as a proxy for 2PL discrimination
parameters. We report clustered bootstrap 95\% confidence intervals
(10{,}000 resamples at the original-item level), analytic standard
errors $\mathrm{SE}(\Delta_{\mathrm{raw}}) = \sigma(\delta_i) / \sqrt{N}$,
and one-sided Wilcoxon signed-rank test $p$-values (testing
$H_1{:}\ \Delta_{\mathrm{raw}} > 0$).


% ==================================================================
\section{Results}
\label{sec:results}

\subsection{Main Results}

Table~\ref{tab:main} presents the empirical contamination
diagnostics for all three evaluated models.

\input{table_main_v3}

\paragraph{All three models show statistically significant
contamination.}
Llama~3.1 8B exhibits the strongest raw contamination signal:\ 92.0\%
accuracy on original GSM8K items versus 50.3\% on $\tau$-isomorphic
variants, yielding $\Delta_{\mathrm{raw}} = +0.417$ (bootstrap 95\%
CI: $[+0.323, +0.506]$, $p < 10^{-10}$). This 41.7-percentage-point
drop is striking given that the variants require \emph{identical
reasoning operations} with only the numeric values changed. Llama~4
Scout 17B shows a comparable pattern ($\Delta_{\mathrm{raw}} = +0.402$,
bootstrap 95\% CI: $[+0.306, +0.498]$, $p < 10^{-10}$), with 94\%
accuracy on originals dropping to 54\% on variants. Items flagged as
memorized ($\delta_i > 0.5$) constitute 43\% of tested items for both
the 8B and Scout models.

\paragraph{Qwen3 confirms contamination at reduced power.}
Qwen3~32B was evaluated on 42 of 100~items due to API daily token
limits, yet still exhibits a highly significant contamination signal:
61.9\% accuracy on originals versus 24.9\% on variants
($\Delta_{\mathrm{raw}} = +0.370$, bootstrap 95\% CI:
$[+0.212, +0.527]$, $p < 10^{-4}$). The extraction-rate gap between
originals and variants is 11.4\% (100\% vs.\ 89\%), which alone
cannot explain the 37.0-point accuracy gap. The contamination signal
is consistent across all three models despite differences in
architecture, parameter count, and training pipeline.

\paragraph{IRT weighting amplifies the contamination signal.}
For all three models, the variance-weighted delta $\Dcont$
substantially exceeds the raw gap: 2.2$\times$ amplification for
Llama~3.1 8B ($\Dcont = +0.915$ vs.\
$\Delta_{\mathrm{raw}} = +0.417$), 2.4$\times$ for Llama~4 Scout
($\Dcont = +0.961$ vs.\ $\Delta_{\mathrm{raw}} = +0.402$), and
2.2$\times$ for Qwen3 ($\Dcont = +0.822$ vs.\
$\Delta_{\mathrm{raw}} = +0.370$). This amplification occurs because
the items with the largest per-item deltas also tend to have high
discrimination (variance between original and variant performance),
receiving greater weight in the IRT-weighted metric.

\paragraph{Per-item analysis reveals bimodal contamination.}
The distribution of per-item deltas $\delta_i$ for each model
exhibits a bimodal pattern: a cluster near $\delta_i = 0$ (items
where the model reasons correctly on both original and variants) and
a heavy right tail at $\delta_i = +1.0$ (items where the model gets
the original correct but fails all variants). This bimodal pattern is
the signature of item-level memorization: certain problems have been
memorized verbatim, while others are solved via genuine reasoning.

\paragraph{Comparison to prior work.}
Our findings are broadly consistent with GSM1K~\cite{zhang2024gsm1k},
which reported 8--13\% performance drops for open-source models on a
parallel benchmark. However, our $\tau$-isomorphic approach reveals
\emph{substantially larger} gaps (up to 41.7\%) for the same model
families, likely because $\tau$-isomorphism provides a stricter
structural guarantee: the variants differ \emph{only} in numeric
values, eliminating the possibility that difficulty differences arise
from surface-level variations in problem framing.

\subsection{Limitations}
\label{sec:limitations}
Several caveats apply to these results. First, the evaluation uses
$T = 1$ trial per item (greedy decoding), preventing estimation of
per-item response variance; future work should use $T \geq 5$ with
temperature $> 0$ to compute proper IRT item parameters. Second,
while the 100-item sample achieves overwhelming statistical
significance for the primary contamination signal ($p < 10^{-10}$),
the full 667-item dataset should be evaluated when API budgets permit
to improve power for detecting small effects and enable
subgroup analyses by reasoning complexity. Third, Qwen3~32B was
evaluated on only 42 of 100~items due to API daily token limits;
ongoing evaluation will complete its assessment. Fourth, the
annotation-chain replay used to compute variant answers cannot handle
problems where the same numeric value serves both as an input and as
an implicit constant (e.g., ``half'' implying division by~2 when the
input value is also~2); such items are identified by the conflation
filter and excluded. Fifth, the IRT-weighted delta uses variance
as a proxy for item discrimination; full 2PL estimation requires
a larger model panel ($\geq 30$ models) than evaluated here.
